\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-19 leslieandkenny.Rnw
Started adding QuasiBinom explanation, made code style more consistent in places

\begin{document}
<<setup, include=FALSE, cache=FALSE,echo=FALSE>>=
require(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, out.width = '\\linewidth',
               out.height = '0.4\\linewidth', dev = 'pdf',
               concordance = TRUE,
               tidy.opts = list(width.cutoff = 60, keep.blank.line=FALSE),
               size = 'footnotesize')
options(replace.assign = TRUE, width = 112, digits = 3, max.print = '202',
        show.signif.stars = FALSE)

@@

# Correctly order nitrogen application levels
fert2$n.trt <- relevel(fert2$n.application, 'FALL')

# If status=INOC is the most interesting case, make INOC the baseline
fert2$status2 <- relevel(fert2$status, 'INOC')

# If YSTN is the variety we want to know about the most, make it baseline
fert2$variety2 <- relevel(fert2$variety, 'YSTN')

@@

<<fakedata,echo=FALSE>>=
# Create some fake data with no differences anywhere
set.seed(983724)
fert.sim1 <- data.frame(
  'block' = c(rep('A', 30), rep('B', 30), rep('C', 30), rep('D', 30), rep('E', 30), rep('F', 30)),
  'variety' = c(rep('Var1', 6), rep('Var2', 6), rep('Var3', 6), rep('Var4', 6), rep('Var5', 6)),
  'n.application' = rep(c('FALL', 'FALL', 'EARLY SPRING', 'EARLY SPRING', 'LATE SPRING', 'LATE SPRING'), 30),
  'status' = rep(c('CNTL', 'INOC'), 90),
  'total' = rep(30, 180),
  'infected' = rnbinom(180, 4, 7/9))
#  'infected' = sample(fert2$infected, 180, replace = TRUE))
@

<<fakepre,echo=FALSE>>=
# Correctly order nitrogen application levels, the default is alphabetical
fert.sim1$n.trt <- relevel(fert.sim1$n.application, 'FALL')

# If status=INOC is the most interesting case, make INOC the baseline
fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')

@@

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\).
\end{itemize}

@@

\begin{figure}[h!]
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = c('boxplot', 'point'), color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

@@

      # Use black for 0, and use heatmap colors (red-orange-yellow-white) for 1 to 30
      col = c('black', heat.colors(30)), zlim = c(0, 30),

      # Don't automatically create axes or labels
      xlab = '', ylab = '', yaxt = 'n', xaxt = 'n')

# Use white line segments to visually separate the varieties
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')

# Place a title at the top, and label Varieties, treatment:status levels, and blocks around the image
title('Infection Counts', line = 4)
axis(3, labels = rep(levels(fert.ordered$block), 5), cex.axis = 0.75,

@@

# Legend
par(mar = c(3, 1, 6, 2))
image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
      col = c('#000000', heat.colors(30)), zlim = c(0, 30))
title('Legend', line = 1.5)
axis(4)
@
\end{figure}

@@

\begin{figure}[h!]
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = variety)
@
\end{figure}

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

@@

<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
@

@@

<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='INOC')

onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
@

@@

Anova(glm3way2, test.statistic = 'F')
Anova(glm2way2, test.statistic = 'F')
anova(glm2way2, glm3way2)
@

@@

noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                    (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS3way)
Anova(noSNMS3way)

noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
                    n.trt*status2 + (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS2way)
Anova(noSNMS2way)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE>>=
par(mfrow = c(2, 2))
plot(resid(glm3way2) ~ predict(glm3way2), col = variety, data = fert2)
plot(resid(glm3way2) ~ fitted(glm3way2), col = variety, data = fert2)
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

@@

<<glmnointeractions, echo=TRUE>>=
glmnointeractions <- glm(cbind(infected, total) ~ block/variety + variety+n.trt
                         + status2,
               family = quasibinomial, data = fert.sim1)
Anova(glmnointeractions, test="F")
@

@@

<<dropindevftest, echo=FALSE>>=
glm.no3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + n.trt*status2 + variety*status2,
               family = quasibinomial, data = fert.sim1)
#drop in dev F-test
#((196-185)/8)/(185/125)
#0.929
#1-pf(0.929, 11, 125)
#0.512
@

@@

\begin{figure}[!h]
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5, fig.align='center'>>=
require(effects)
plot(Effect(c('variety','n.trt','status2'), glm3way),
     multiline = TRUE, type = 'response', se = TRUE, ci.style='bars',
     x.var = 'n.trt', rotx = 45, layout = c(5, 1),
     main = 'Variety/Treatment/Inoculation Effects Plot',
     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
     key.args = list(title = 'Inoculation Status'))
@
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
\end{figure}

@@

<<summary, echo=TRUE>>=
Effect(c('variety','n.trt','status2'), glm3way, se=TRUE)
@

>>>>

\begin{document}
<<setup, include=FALSE, cache=FALSE,echo=FALSE>>=
require(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, out.width = '\\linewidth',
               out.height = '0.4\\linewidth', dev = 'pdf',
               concordance = TRUE,
               tidy.opts = list(width.cutoff = 60, keep.blank.line=FALSE),
               size = 'footnotesize')
options(replace.assign = TRUE, width = 112, digits = 3, max.print = '202',
        show.signif.stars = FALSE)
knit_theme$set('print')

@@

# Correctly order nitrogen application levels
fert2$n.trt <- relevel(fert2$n.application, "FALL")

# If status=INOC is the most interesting case, make INOC the baseline
fert2$status2 <- relevel(fert2$status, "INOC")

# If YSTN is the variety we want to know about the most, make it baseline
fert2$variety2 <- relevel(fert2$variety, "YSTN")

@@

<<fakedata,echo=FALSE>>=
# Create some fake data with no differences anywhere
set.seed(983724)
fert.sim1 <- data.frame(expand.grid(
  "status" = c("CNTL", "INOC"),
  "n.application" = c("FALL", "EARLY SPRING", "LATE SPRING"),
  "variety" = c("Var1", "Var2", "Var3", "Var4", "Var5"),
  "block" = LETTERS[1:6],
  "total" = 30),
  "infected" = rnbinom(180, 4, 7/9))
#  "infected" = sample(fert2$infected, 180, replace = TRUE))
@

<<fakepre,echo=FALSE>>=
# Correctly order nitrogen application levels, the default is alphabetical
fert.sim1$n.trt <- relevel(fert.sim1$n.application, "FALL")

# If status=INOC is the most interesting case, make INOC the baseline
fert.sim1$status2 <- relevel(fert.sim1$status, "INOC")

@@

\noindent The model is
$$
y_i\sim\mathrm{overdispersed Binomial}(30, p_i, \omega),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\omega\) is an overdispersion parameter, so that the standard
deviation of the response in plot \(i\) is \(\sqrt{30\omega p_i(1-p_i)}\).
\end{itemize}

@@

\begin{figure}[h!]
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = c("boxplot", "point"), color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

@@

      # Use black for 0, and use heatmap colors (red-orange-yellow-white) for 1 to 30
      col = c("black", heat.colors(30)), zlim = c(0, 30),

      # Don't automatically create axes or labels
      xlab = "", ylab = "", yaxt = "n", xaxt = "n")

# Use white line segments to visually separate the varieties
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = "white")

# Place a title at the top, and label Varieties, treatment:status levels, and blocks around the image
title("Infection Counts", line = 4)
axis(3, labels = rep(levels(fert.ordered$block), 5), cex.axis = 0.75,

@@

# Legend
par(mar = c(3, 1, 6, 2))
image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = "",
      col = c("#000000", heat.colors(30)), zlim = c(0, 30))
title("Legend", line = 1.5)
axis(4)
@
\end{figure}

@@

\begin{figure}[h!]
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = "variety")
@
\end{figure}

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

@@

<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
              family = quasibinomial, data = fert.sim1, subset = variety!="Var1")
@

@@

<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=="INOC")

onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=="CNTL")
@

@@

Anova(glm3way2, test.statistic = "F")
Anova(glm2way2, test.statistic = "F")
anova(glm2way2, glm3way2)
@

@@

noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                    (1|block/row), subset = variety != "SNMS",
                    control = glmerControl(optimizer = "bobyqa"),
                    family = binomial, data = fert2)
summary(noSNMS3way)
Anova(noSNMS3way)

noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
                    n.trt*status2 + (1|block/row), subset = variety != "SNMS",
                    control = glmerControl(optimizer = "bobyqa"),
                    family = binomial, data = fert2)
summary(noSNMS2way)
Anova(noSNMS2way)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE>>=
par(mfrow = c(2, 2))
plot(resid(glm3way2) ~ predict(glm3way2), col = variety, data = fert2)
plot(resid(glm3way2) ~ fitted(glm3way2), col = variety, data = fert2)
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!="SNMS")
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!="SNMS")
legend("topright", pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

@@

<<glmnointeractions, echo=TRUE>>=
glmnointeractions <- glm(cbind(infected, total) ~ block*variety + variety + n.trt + status2,
               family = quasibinomial, data = fert.sim1)
Anova(glmnointeractions, test="F")
@

@@

<<dropindevftest, echo=FALSE>>=
glm.no3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt + n.trt*status2 + variety*status2,
               family = quasibinomial, data = fert.sim1)
#drop in dev F-test
#((196-185)/8)/(185/125)
#0.929
#1-pf(0.929, 11, 125)
#0.512
@

@@

\begin{figure}[!h]
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5, fig.align='center'>>=
require(effects)
plot(Effect(c("variety", "n.trt", "status2"), glm3way),
     multiline = TRUE, type = "response", se = TRUE, ci.style="bars",
     x.var = "n.trt", rotx = 45, layout = c(5, 1),
     main = "Variety/Treatment/Inoculation Effects Plot",
     xlab = "Nitrogen Application Date", ylab = "Infection Probability",
     key.args = list(title = "Inoculation Status"))
@
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
\end{figure}

@@

<<summary, echo=TRUE>>=
Effect(c("variety", "n.trt", "status2"), glm3way, se=TRUE)
@
